---
title: "spectral_explore"
author: "Holden Jones"
date: "2026-04-22"
output: html_document
---

Begin exploring with Eben Broadbent's data - this is preprocessed.

load packages
```{r}
library(terra)
library(biodivMapR)
```

define input and output paths
- note that using external harddrive for data storage
- b/c massive datasizes
```{r}
# ------------------------------------------------------------
# 1. DEFINE PATHS
# ------------------------------------------------------------

# Directory containing your individual band TIFs
ortho_dir <- '/Volumes/SHJ_Drive2/gator_eye/072022/G2 - Las Cruces/Multispectral_ortho/Orthomosaic'

# Base name prefix (everything before the band suffix)
prefix <- 'GatorEye_SubjectTerms_f072022_G2_p120422_v1_transparent_mosaic'

# Individual band file paths
blue_path     <- file.path(ortho_dir, paste0(prefix, '_blue.tif'))
green_path    <- file.path(ortho_dir, paste0(prefix, '_green.tif'))
red_path      <- file.path(ortho_dir, paste0(prefix, '_red.tif'))
rededge_path  <- file.path(ortho_dir, paste0(prefix, '_red edge.tif'))
nir_path      <- file.path(ortho_dir, paste0(prefix, '_nir.tif'))
# NOTE: _lwir.tif (thermal) is intentionally excluded

# Output directory
output_dir <- '/Volumes/SHJ_Drive2/gator_eye/072022/G2 - Las Cruces/Multispectral_ortho/Orthomosaic/biodivMapR_results'
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

stacked_path <- file.path(output_dir, 'altumPT_5band_stack.tif')
```

checking how many layers actually in single-band TIFS
```{r}
# Check how many bands (layers) each file actually has
cat('Blue bands:   ', terra::nlyr(terra::rast(blue_path)), '\n')
cat('Green bands:  ', terra::nlyr(terra::rast(green_path)), '\n')
cat('Red bands:    ', terra::nlyr(terra::rast(red_path)), '\n')
cat('RedEdge bands:', terra::nlyr(terra::rast(rededge_path)), '\n')
cat('NIR bands:    ', terra::nlyr(terra::rast(nir_path)), '\n')
```

they have 2 - second is an alpha transparency channel
- now stack - this will write a large file, so do a sanity check first
```{r}
# Select only band 1 (reflectance) from each file, dropping the alpha channel (band 2)
# stack <- c(
#   terra::rast(blue_path)[[1]],
#   terra::rast(green_path)[[1]],
#   terra::rast(red_path)[[1]],
#   terra::rast(rededge_path)[[1]],
#   terra::rast(nir_path)[[1]]
# )
# names(stack) <- c('Blue_475', 'Green_560', 'Red_668', 'RedEdge_717', 'NIR_842')
# 
# terra::writeRaster(stack, stacked_path, overwrite = TRUE, datatype = 'FLT4S')
# cat('Stack saved to:', stacked_path, '\n')
```

ensure that the individual layers are ready for stacking
```{r}
# Quick sanity check before writing
# print(stack)          # confirms 5 layers, correct names, CRS, extent
# terra::nlyr(stack)    # should return 5
```

stack them here
```{r}
# comment out b/c this takes a really long time
# terra::writeRaster(stack, stacked_path, overwrite = TRUE, datatype = 'FLT4S')
# cat('Stack saved to:', stacked_path, '\n')
```

run reflectance scale check
```{r}
# stack <- terra::rast(stacked_path)
# print(terra::global(stack, fun = 'range', na.rm = TRUE))
```

create vegetation mask - important to remove non veg from spectral diversity estimates
```{r}
# ------------------------------------------------------------
# 4. BUILD VEGETATION MASK
#    Uses NDVI (NIR + Red) and NIR threshold to exclude:
#    - bare soil, water, impervious surfaces (low NDVI)
#    - deep shadows (low NIR)
#    Adjust thresholds based on your ecosystem and reflectance scale
# ------------------------------------------------------------

mask_path <- file.path(output_dir, 'vegetation_mask.tif')

if (!file.exists(mask_path)) {
  cat('\nBuilding vegetation mask...\n')
  
  Red <- stack[[3]]
  NIR <- stack[[5]]
  
  NDVI <- (NIR - Red) / (NIR + Red)
  
  # ---- TUNE THESE THRESHOLDS ----
  # Data is in raw DN scale (0-65535), so NIR threshold is scaled accordingly.
  # NDVI is always in [-1, 1] regardless of input scale.
  NDVI_thresh <- 0.3      # exclude NDVI < 0.3 (non-vegetation)
  NIR_thresh  <- 6554     # ~10% of 65535 scale; exclude deep shadows
  # --------------------------------
  
  # Use terra::ifel() instead of as.integer() — base R coercion fails on SpatRaster
  mask <- terra::ifel((NDVI >= NDVI_thresh) & (NIR >= NIR_thresh), 1L, 0L)
  
  terra::writeRaster(mask, mask_path, overwrite = TRUE, datatype = 'INT1U')
  cat('Mask saved. Check it in QGIS before proceeding!\n')
}

# IMPORTANT: Open vegetation_mask.tif in QGIS and visually verify 
# that vegetation is 1 (white) and non-vegetation/shadows are 0 (black)
# Adjust NDVI_thresh and NIR_thresh above if needed, then re-run
```

vegetation mask looks fine - probably would want to adjust the thresholds if 
using fr but as exploratory it checks out

radiometric filtering - another step for refining the mask?
- in this case there's no difference compared to the manual vegetation_mask
- but sometimes this will catch more issues in the blue band
```{r}
# ------------------------------------------------------------
# 5. RADIOMETRIC FILTERING (biodivMapR refines mask further)
# ------------------------------------------------------------

sensor_bands <- c(475, 560, 668, 717, 842)  # Altum-PT center wavelengths (nm)

output_dir_PCA <- file.path(output_dir, 'SPCA')
dir.create(output_dir_PCA, showWarnings = FALSE, recursive = TRUE)

# Thresholds scaled to raw DN range (0-65535).
# NDVI is unitless [-1, 1] and needs no scaling.
# NIR and Blue thresholds are ~10% and ~15% of 65535 respectively.
mask_path_PCA <- radiometric_filtering(
  input_raster_path = stacked_path,
  input_rast_wl     = sensor_bands,
  input_mask_path   = mask_path,
  output_dir        = output_dir_PCA,
  NDVI_Thresh       = 0.3,
  Blue_Thresh       = 9830,
  NIR_Thresh        = 6554
)
```








## BELOW IS THE CODE PRODUCED BY CLAUDE - ABOVE INTEGRATE IN SMALLER CHUNKS

```{r}
# ============================================================
# biodivMapR Workflow: MicaSense Altum-PT (Pix4D single-band TIFs)
# ============================================================

library(terra)
library(biodivMapR)

# ------------------------------------------------------------
# 1. DEFINE PATHS
# ------------------------------------------------------------

# Directory containing your individual band TIFs
ortho_dir <- '/data/Orthomosaic'

# Base name prefix (everything before the band suffix)
prefix <- 'GatorEye_SubjectTerms_f072022_G2_p120422_v1_transparent_mosaic'

# Individual band file paths
blue_path     <- file.path(ortho_dir, paste0(prefix, '_blue.tif'))
green_path    <- file.path(ortho_dir, paste0(prefix, '_green.tif'))
red_path      <- file.path(ortho_dir, paste0(prefix, '_red.tif'))
rededge_path  <- file.path(ortho_dir, paste0(prefix, '_red edge.tif'))
nir_path      <- file.path(ortho_dir, paste0(prefix, '_nir.tif'))
# NOTE: _lwir.tif (thermal) is intentionally excluded

# Output directory
output_dir <- '/path/to/072022/G2 - Las Cruces/Multispectral_ortho/biodivMapR_results'
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# ------------------------------------------------------------
# 2. STACK BANDS INTO A SINGLE 5-BAND GEOTIFF
#    Band order: Blue, Green, Red, RedEdge, NIR
# ------------------------------------------------------------

stacked_path <- file.path(output_dir, 'altumPT_5band_stack.tif')

if (!file.exists(stacked_path)) {
  cat('Stacking individual band TIFs...\n')
  
  blue     <- terra::rast(blue_path)
  green    <- terra::rast(green_path)
  red      <- terra::rast(red_path)
  rededge  <- terra::rast(rededge_path)
  nir      <- terra::rast(nir_path)
  
  # Stack in order: Blue=1, Green=2, Red=3, RedEdge=4, NIR=5
  stack <- c(blue, green, red, rededge, nir)
  names(stack) <- c('Blue_475', 'Green_560', 'Red_668', 'RedEdge_717', 'NIR_842')
  
  terra::writeRaster(stack, stacked_path, overwrite = TRUE,
                     datatype = 'FLT4S')  # float preserves reflectance values
  cat('Stack saved to:', stacked_path, '\n')
} else {
  cat('Stack already exists, loading...\n')
}

stack <- terra::rast(stacked_path)

# ------------------------------------------------------------
# 3. CHECK REFLECTANCE SCALE
#    Pix4D transparent_mosaic outputs are typically in 0-1 or 0-65535
#    biodivMapR expects values in a consistent numeric range
#    Print a quick summary to confirm
# ------------------------------------------------------------

cat('\n--- Reflectance value check (sample of band statistics) ---\n')
print(terra::global(stack, fun = 'range', na.rm = TRUE))

# If values are 0-65535 (scaled integer), normalize to 0-1:
# stack <- stack / 65535
# If values are 0-1 already, proceed as-is.
# If values are 0-10000 (Pix4D scaled reflectance), that is also fine — 
# just keep Blue_Thresh etc. consistent with that scale in Step 5.

rast_path <- stacked_path  # used in subsequent steps

# ------------------------------------------------------------
# 4. BUILD VEGETATION MASK
#    Uses NDVI (NIR + Red) and NIR threshold to exclude:
#    - bare soil, water, impervious surfaces (low NDVI)
#    - deep shadows (low NIR)
#    Adjust thresholds based on your ecosystem and reflectance scale
# ------------------------------------------------------------

mask_path <- file.path(output_dir, 'vegetation_mask.tif')

if (!file.exists(mask_path)) {
  cat('\nBuilding vegetation mask...\n')
  
  Red <- stack[[3]]
  NIR <- stack[[5]]
  
  NDVI <- (NIR - Red) / (NIR + Red)
  
  # ---- TUNE THESE THRESHOLDS ----
  # For reflectance scale 0-1:
  NDVI_thresh <- 0.3    # exclude NDVI < 0.3 (non-vegetation)
  NIR_thresh  <- 0.1    # exclude NIR < 0.1 (shadows)
  # For scale 0-10000, multiply thresholds by 10000
  # --------------------------------
  
  mask <- as.integer((NDVI >= NDVI_thresh) & (NIR >= NIR_thresh))
  
  terra::writeRaster(mask, mask_path, overwrite = TRUE, datatype = 'INT1U')
  cat('Mask saved. Check it in QGIS before proceeding!\n')
}

# IMPORTANT: Open vegetation_mask.tif in QGIS and visually verify 
# that vegetation is 1 (white) and non-vegetation/shadows are 0 (black)
# Adjust NDVI_thresh and NIR_thresh above if needed, then re-run

# ------------------------------------------------------------
# 5. RADIOMETRIC FILTERING (biodivMapR refines mask further)
# ------------------------------------------------------------

sensor_bands <- c(475, 560, 668, 717, 842)  # Altum-PT center wavelengths (nm)

output_dir_PCA <- file.path(output_dir, 'SPCA')
dir.create(output_dir_PCA, showWarnings = FALSE, recursive = TRUE)

# Adjust thresholds to match your reflectance scale!
# For 0-1 scale:    NDVI=0.3, Blue=0.15, NIR=0.1
# For 0-10000 scale: NDVI=0.3, Blue=1500, NIR=1000
mask_path_PCA <- radiometric_filtering(
  input_raster_path = rast_path,
  input_rast_wl     = sensor_bands,
  input_mask_path   = mask_path,
  output_dir        = output_dir_PCA,
  NDVI_Thresh       = 0.3,
  Blue_Thresh       = 0.15,
  NIR_Thresh        = 0.1
)

# ------------------------------------------------------------
# 6. SPECTRAL TRANSFORMATION (SPCA with Continuum Removal)
#    Continuum Removal is recommended for high-res drone data
#    to reduce within-crown illumination effects
# ------------------------------------------------------------

cat('\nRunning SPCA...\n')
PCA_Output <- perform_PCA(
  input_raster_path = rast_path,
  input_rast_wl     = sensor_bands,
  output_dir        = output_dir_PCA,
  Continuum_Removal = TRUE,
  input_mask_path   = mask_path_PCA,
  maxRows           = 1000
)

# ------------------------------------------------------------
# 7. VISUALLY INSPECT PCA COMPONENTS
#    Open the PCA output in QGIS or plot here
#    Select components that show meaningful vegetation patterns
#    (vegetation type boundaries, not noise or striping artifacts)
# ------------------------------------------------------------

pca_rast <- terra::rast(PCA_Output$PCA_Files$PCA)
cat('\nNumber of PCA components:', terra::nlyr(pca_rast), '\n')

# Plot all components for visual inspection
terra::plot(pca_rast, main = paste0('PC', 1:terra::nlyr(pca_rast)))

# Update selected_bands after visual inspection:
selected_bands <- c(1, 2, 3, 4)  # typically skip the last noisy PC

# ------------------------------------------------------------
# 8. RUN biodivMapR — α AND β DIVERSITY
# ------------------------------------------------------------

output_dir_biodivMapR <- file.path(output_dir, 'biodivMapR_SPCA')
dir.create(output_dir_biodivMapR, showWarnings = FALSE, recursive = TRUE)

Kmeans_info_save <- file.path(output_dir_biodivMapR, 'Kmeans_info.RData')
Beta_info_save   <- file.path(output_dir_biodivMapR, 'Beta_info.RData')

options_bdmr <- set_options_biodivMapR(fun = 'biodivMapR_full')
options_bdmr$nb_clusters   <- 50   # number of spectral species
options_bdmr$maxRows       <- 1000
options_bdmr$alpha_metrics <- c('richness', 'shannon', 'hill')

# --- WINDOW SIZE GUIDE for Altum-PT ---
# At 120m AGL: ~5.3 cm/px  →  window=10 ≈ 0.53m, window=30 ≈ 1.6m, window=100 ≈ 5.3m
# At  60m AGL: ~2.6 cm/px  →  window=10 ≈ 0.26m, window=50 ≈ 1.3m, window=100 ≈ 2.6m
# Match window to the scale of your ecological unit of interest (e.g. tree crown, plot)
window_size <- 30  # adjust based on your flight altitude and ecological scale

cat('\nRunning biodivMapR (this may take a while for large orthomosaics)...\n')
ab_info <- biodivMapR_full(
  input_raster_path = PCA_Output$PCA_Files$PCA,
  input_mask_path   = mask_path_PCA,
  output_dir        = output_dir_biodivMapR,
  selected_bands    = selected_bands,
  window_size       = window_size,
  Kmeans_info_save  = Kmeans_info_save,
  Beta_info_save    = Beta_info_save,
  nbCPU             = 4,           # set to your available CPU cores
  options           = options_bdmr
)

cat('\nDone! Output diversity maps are in:\n', output_dir_biodivMapR, '\n')
cat('Load them in QGIS or R with terra::rast() for visualization.\n')
```









